---
title: "R Notebook"
output: html_notebook
---

How often people rebuy ?
My goal is to take a look on the customer behavior and find if we can cluster them by how often they buy.
The steps that I'll take:
Group the data by users
How many users haven't bought again ?
Histogram
Cluster

Group the data by Users

```{r}
options(warn=-1)
library(ggplot2) 
library(readr) 
library(dplyr)
library(ggplot2)

orders <- read_csv("../input/orders.csv")

user_freq <- orders %>%
    select(user_id, order_number, days_since_prior_order) %>%
    group_by(user_id) %>%
    summarise(total = max(order_number), 
              frequency = mean(days_since_prior_order, na.rm = TRUE))


glimpse(user_freq)
```
How many users haven't bought again ?
I didn't know this, but looks like we just have users that have bought more than once.
This means that we don't have to clean it.
Take a look :
```{r}
user_freq %>%
    filter(total == 1) %>%
    glimpse
```

Histogram
1 - How many days there are between orders :

```{r}
ggplot(user_freq, aes(frequency)) +
    geom_histogram()
```

2 - Number of orders :
In [4]:
```{r}
ggplot(user_freq, aes(total)) +
    geom_histogram()
```

Cluster
```{r}
set.seed(42)
clus <- kmeans(user_freq[,2:3], 4)
clus$cluster <- as.factor(clus$cluster)
(clus$centers)
```
Cluster results looks very good
We can see that there are some type of users, and the types are:
Buy twice a week
Buy almost monthly
Buy almost weekly
Buy almost every two weeks
```{r}
ggplot(user_freq,
       aes(total, frequency, color = clus$cluster)) +
    geom_point()
```

